# Factor Analysis and Principal Component Analysis {#factor-analysis-PCA}

## Getting Started {#gettingStarted-factorAnalysis}

### Load Libraries {#loadLibraries-factorAnalysis}

```{r}
library("petersenlab")
library("lavaan")
library("psych")
library("corrplot")
library("nFactors")
library("semPlot")
library("lavaan")
library("semTools")
library("dagitty")
library("kableExtra")
library("MOTE")
library("tidyverse")
library("here")
library("tinytex")
library("knitr")
library("rmarkdown")
```

### Load Data {#loadData-factorAnalysis}

### Prepare Data {#prepareData-factorAnalysis}

#### Add Missing Data {#addMissingData-factorAnalysis}

Adding missing data to dataframes helps make examples more realistic to real-life data and helps you get in the habit of programming to account for missing data.
For reproducibility, I set the seed below.
Using the same seed will yield the same answer every time.
There is nothing special about this particular seed.

`HolzingerSwineford1939` is a data set from the `lavaan` package [@R-lavaan] that contains mental ability test scores (`x1`–`x9`) for seventh- and eighth-grade children.

```{r}
set.seed(52242)

varNames <- names(HolzingerSwineford1939)
dimensionsDf <- dim(HolzingerSwineford1939)
unlistedDf <- unlist(HolzingerSwineford1939)
unlistedDf[sample(
  1:length(unlistedDf),
  size = .15 * length(unlistedDf))] <- NA
HolzingerSwineford1939 <- as.data.frame(matrix(
  unlistedDf,
  ncol = dimensionsDf[2]))
names(HolzingerSwineford1939) <- varNames
vars <- c("x1","x2","x3","x4","x5","x6","x7","x8","x9")
```
